### 2018-09-17 Manuscripts/cog-psych/revision/cogpsych.tex
Re manuscript: some new paragraphs (not yet vetted by Joe) on real world example and discussion on possible influences of contrast. Re: new exp, individual plots and some scripts to make life a little easier.

To successfully generate something novel, what is generated must be different
from what is already known. This fundamental constraint, ``being different'', or
contrasting from other categories in the relevant domain, is the focus of our
work. Although implicitly assumed in some work, this constraint has been
overlooked in previous research: To our knowledge, there has not been any
systematic investigation addressing how generated categories {\em differ} from
what is already known. Although the idea of category contrast is ubiquitous
throughout the categorization literature, and extends to a variety of other
fields \citep[e.g., color;][]{regier2007}, the idea that a new category should
be ``different'' is vague, as there are many ways it could be different from a
previously observed category.

@@

In our simulations, we set the prior mean $\mu_0$ to the center of the stimulus
space the prior variance to be isotropic ($\Sigma_0 = \lambda \mathbf{I}$ where
$\lambda$ is a free parameter and $\mathbf{I}$ is a $d$-by-$d$ identity matrix
with $d$ representing the number of dimensions or features in the domain). The
$\kappa$ and $\nu$ parameters are freely estimated within the constraints
$\kappa > d-1$ and $\nu > 0$.

@@

The broader study of creativity currently involves a wide breadth of different
approaches,\citep[for a review see][]{kozbelt2010theories}, such as those based
on free association \citep{mednick1962associative} and conceptual combination
\citep{estes2002emergence,murphy1988comprehending}. In addition, recent work in
the machine learning literature has explored using neural networks to address
the overall problem of creative generation
\citep[e.g.,][]{goodfellow2014gan,ho2016gail,kingma2016,chen2016infogan}. In
contrast to these varied investigations, we reduced our focus by studying a
highly complex behavior (category generation) as it applies within a
well-established domain (artificial category learning). We hope that future work
incorporates and highlights the importance of contrast into theories of
creativity.

@@

Interestingly, the unidimensional sort bias that is commonly observed in
unsupervised cateogorization
\citep{imai1965discriminability,milton2004influence,ahn1992two}, where
participants sort unlabelled family resemblance category exemplars by focusing
on a single feature, was not consistently observed in our tasks. In our
experiments we only observed this bias in the row condition of Experiment 1,
whereas in other conditions participants generated categories that typically
varied along both dimensions. This may be unsurprising at face value, since
participants in conditions other than the row condition were trained on a prior
category that varied on both dimensions. However, these results suggest that the
unidimensional sort bias is not universal to all types of category construction
-- or more specifically, the unidimensional bias may not apply when participants
need to generate categories with entirely new exemplars.

@@

While the contrast models were born quite naturally out of the corresponding
classical models, it is worth considering if similar benefits in prediction
would be seen if the contrast mechanisms were applied to different classical
models. Although there is no clear way to adapt the exemplar dissimilarity
mechanism to the current hierarchical Bayesian model, it is a fairly
straightforward process applying the representativeness mechanism to an exemplar
model. Specifically, by treating the similarity measure (Equation
\ref{eq:similarity}) in the current implementation of copy-and-tweak as a
density estimate for the representativeness mechanism (Equation
\ref{eq:representativeness}), we gain a new model of representativeness that is
based on exemplar similarity instead of a multivariate Gaussian
likelihood.\footnote{An analogous implementation of the representativeness
  mechanism to PACKER results in both $\theta_c$ and $\theta_t$ adding to
  constant value that is independent of the similarity to contrast and target
  exemplars respectively. Consequently, it becomes equivalent to a
  copy-and-tweak model with representativeness.}

@@

>>>>

To successfully generate something novel, what is generated must be different
from what is already known. This fundamental constraint, ``being different'', or
contrasting from other categories in the relevant domain, is the focus of our
work. Intuitively, the influence of contrast may be expected in a number of
real-life scenarios involving the generation of new categories. For instance, a
college student who has been eating instant noodles for a week may be more
inclined to seek out something healthier and less carbohydrate-heavy (e.g., a
Caesar salad) compared to something similar (a different flavour of instant
noodles). An actor who is dissatisfied with being typecast as a villain may seek
out roles as heroes as opposed to playing other antagonistic characters. A more
compelling example is seen in recent evidence by \cite{askin2017} who found that
new songs that were more different to their peers were generally more popular
than new songs that were similar to others. It appears that once there is the
motivation to generate something new, there is an associated motivation to
generate something \emph{different}.

Although implicitly assumed in some work, this constraint has been
overlooked in previous research: To our knowledge, there has not been any
systematic investigation addressing how generated categories {\em differ} from
what is already known. Although the idea of category contrast is ubiquitous
throughout the categorization literature, and extends to a variety of other
fields \citep[e.g., color;][]{regier2007}, the idea that a new category should
be ``different'' is vague, as there are many ways it could be different from a
previously observed category.

@@

In our simulations, we set the prior mean $\mu_0$ to the center of the stimulus
space and the prior variance to be isotropic ($\Sigma_0 = \lambda \mathbf{I}$ where
$\lambda$ is a free parameter and $\mathbf{I}$ is a $d$-by-$d$ identity matrix
with $d$ representing the number of dimensions or features in the domain). The
$\kappa$ and $\nu$ parameters are freely estimated within the constraints
$\kappa > d-1$ and $\nu > 0$.

@@

The broader study of creativity currently involves a wide breadth of different
approaches\citep[for a review see][]{kozbelt2010theories}, such as those based
on free association \citep{mednick1962associative} and conceptual combination
\citep{estes2002emergence,murphy1988comprehending}. In addition, recent work in
the machine learning literature has explored using neural networks to address
the overall problem of creative generation
\citep[e.g.,][]{goodfellow2014gan,ho2016gail,kingma2016,chen2016infogan}. In
contrast to these varied investigations, we reduced our focus by studying a
highly complex behavior (category generation) as it applies within a
well-established domain (artificial category learning). We hope that future work
incorporates and highlights the importance of contrast into theories of
creativity.

@@

Interestingly, the unidimensional sort bias that is commonly observed in
unsupervised categorization
\citep{imai1965discriminability,milton2004influence,ahn1992two}, where
participants sort unlabelled family resemblance category exemplars by focusing
on a single feature, was not consistently observed in our tasks. In our
experiments we only observed this bias in the row condition of Experiment 1,
whereas in other conditions participants generated categories that typically
varied along both dimensions. This may be unsurprising at face value, since
participants in conditions other than the row condition were trained on a prior
category that varied on both dimensions. However, these results suggest that the
unidimensional sort bias is not universal to all types of category construction
-- or more specifically, the unidimensional bias may not apply when participants
need to generate categories with entirely new exemplars.

@@

While the contrast models were born quite naturally out of the corresponding
classical models, it is worth considering if similar benefits in prediction
would be seen if the contrast mechanisms were applied to different classical
models. Although there is no clear way to adapt the exemplar dissimilarity
mechanism to the current hierarchical Bayesian model, it is a fairly
straightforward process applying the representativeness mechanism to an exemplar
model. Specifically, by treating the similarity measure (Equation
\ref{eq:similarity}) in the current implementation of copy-and-tweak as a
density estimate for the representativeness mechanism (Equation
\ref{eq:representativeness}), we gain a new model of representativeness that is
based on exemplar similarity instead of a multivariate Gaussian
likelihood.\footnote{An analogous implementation of the representativeness
  mechanism to PACKER results in both $\theta_c$ and $\theta_t$ adding to
  constant value that is independent of the similarity to contrast and target
  exemplars respectively. Consequently, it is formally equivalent to a
  copy-and-tweak model with representativeness.}

@@

The present work focuses on the influence of contrast on categorization,
primarily by exploring two different conceptualizations of contrast. However,
throughout this paper we have only explored one type of category generation
problem, that is the generation of new categories of squares that vary only in
size and shade. It is possible that the strength of the influence of contrast on
category generation decisions varies depending on the nature of the category
generation problem, for instance, in situations where the boundaries of the
domain are not clearly defined. To provide some intuition for this, consider the
following example: an entomologist is asked to draw a new type of insect that
they have never seen. Clearly, features of the new insect will be constrained by
the definition of an insect: it will be an arthropod with six legs and a
relatively hard exoskeleton. However, there are no strict limitations on other
features such as how large the insect needs to be, or if the insect should have
a particular type of pattern on its body. In this scenario, it is plausible that
the entomologist will draw an insect that is fairly similar to insects that
currently exist (e.g., small in size with a camouflage pattern that matches its
environment) as opposed to something completely different (e.g., a horse-sized
beetle with roman numerals on its back), indicating that contrast plays only a
minor role in the decision making process. In this situation it appears that
there is a combination of clearly defined constraints on the domain boundary
that are imposed by the external environment (insects must be arthropods with
six legs) but also ill-defined constraints that are imposed by the observer
(insects are probably small and have patterns on their body that suit the
environment). However, the relation of these contraints (and the interactions
between them) to the role of contrast is not currently well understood.

The definition of the domain boundary is only one possible factor that could
influence the effect of contrast on categorization. It is also plausible that
other conditions such as the nature of the instruction (``Generate a new
category that is Not Alpha'' vs ``Generate a new category that is Beta'') and
the boundedness of the features (using features such as hue and orientation that
do have have strict minima or maxima) can affect the extent to which contrast
influences categorization. Though our current contrast models can allow the
influence of contrast to vary, they do not make any predictions regarding the
specific conditions under which contrast can vary. Consequently,
addressing this issue would be an immensely promising avenue for future
research.
